\chapter{System Infrastructure}

\section{Groups of Processes}
\label{sec:pid_groups}
\erlmoduleindex{pid\_groups}

\begin{itemize}
\item What is it? How to distinguish from Erlangs internal named processes?
\item Joining a process group
\item Why do we do this... (managing several independent nodes inside a single
  Erlang VM for testing)
\end{itemize}

\section{\texorpdfstring{The Communication Layer \erlmodule{comm}}
          {The Communication Layer comm}}
\label{sec:comm}
\erlmoduleindex{comm}

\begin{itemize}
\item in general
\item format of messages (tuples)
\item use messages with cookies (server and client side)
\item What is a message tag?
\end{itemize}

\input{gen_component}

\section{The Process' Database (\texttt{pdb})}
\erlmoduleindex{pdb}

\begin{itemize}
\item How to use it and how to switch from erlang:put/set to ets and implied
  limitations.
\end{itemize}

\section{Failure Detectors (\texttt{fd})}
\erlmoduleindex{fd}

\begin{itemize}
\item uses Erlang monitors locally
\item is independent of component load
\item uses heartbeats between Erlang virtual machines
\item uses a single proxy heartbeat server per Erlang virtual machine, which
  itself uses Erlang monitors to monitor locally
\item uses dynamic timeouts to implement an eventually perfect failure detector.
\end{itemize}

\section{Monitoring Statistics (\texttt{monitor}, \texttt{rrd})}
\erlmoduleindex{monitor}

The \erlmodule{monitor} module offers several methods to gather meaningful
statistics using the \code{rrd()} data type defined in \erlmodule{rrd}.

\erlmoduleindex{rrd}
\code{rrd()} records work with time slots, i.e. a fixed slot length is given
at creation and items which should be inserted will be either put into the
current slot, or a new slot will be created.
Each data item thus needs a time stamp associated with it. It must not be a
real time, but can also be a virtual time stamp.

The \erlmodule{rrd} module thus offers two different APIs:
one with transparent time handling, e.g.
 \erlfun{rrd}{create}{/3}, \erlfun{rrd}{add\_now}{/2}, and
one with manual time handling, e.g.
 \erlfun{rrd}{create}{/4}, \erlfun{rrd}{add}{/3}.

To allow different evaluations of the stored data, the following types of data
are supported:
\begin{itemize}
  \item \code{gauge}: only stores the newest value of a time slot, e.g. for thermometers,
  \item \code{counter}: sums up all values inside a time slot,
  \item \code{timing}: records time spans and stores values to easily calculate e.g. the sum, the
  standard deviation, the number of events, the min and max,
  \item \code{timing_with_hist}: similar to \code{timing} but also records a more detailed
  (approximated) histogram of the data,
  \item \code{event}: records each event (including its time stamp) inside a time slot in a list
  (this should be rarely used as the amount of data stored may be very big).
  \item \code{histogram, N}: records values in an approximative histogram of size N
  \item \code{histogram_rt, N, BaseKey}: histogram of size N which
    operates on the key space of the DHT. BaseKey is the key with the
    largest distance to all keys in the histogram.
\end{itemize}

The \erlmodule{monitor} offers functions to conveniently store and retrieve
such values. It is also started as a process in each \code{dht_node} and
\code{basic_services} group as well as inside each \code{clients_group}.
This process ultimately stores the whole rrd() structure.
There are three paradigms how values can be stored:
\begin{enumerate}
  \item Values are gathered in the process that is generating the values.
        Inside this process, the \code{rrd()} is stored in the erlang
        dictionary.
        Whenever a new time slot is started, the values will be
        reported to the monitor process of the gathering process' group.
  \item Values are gathered in the process that is generating the values.
        Inside this process, the \code{rrd()} is handled manually.
        After changing the \code{rrd()}, a manual check for reporting needs to
        be issued using \erlfun{monitor}{check\_report}{/4}.
  \item Values are immediately send to the monitor process where it undergoes
        the same procedures until it is finally stored and available to other
        processes. This is especially useful if the process generating the
        values does not live long or does not regularly create new data, e.g.
        the client.
\end{enumerate}

The following example illustrates the first mode, i.e. gathering data in the
generating process. It has been taken from the \erlmodule{cyclon} module which
uses a \code{counter} data type:

\begin{lstlisting}
% initialise the monitor with an empty rrd() using a 60s monitoring interval
monitor:proc_set_value(?MODULE, 'shuffle', rrd:create(60 * 1000000, 3, counter)),
% update the value by adding one
monitor:proc_set_value(?MODULE, 'shuffle', fun(Old) -> rrd:add_now(1, Old) end),
% check regularly whether to report the data to the monitor:
monitor:proc_check_timeslot(?MODULE, 'shuffle')
\end{lstlisting}

The first two parameters of \erlfun{monitor}{proc\_set\_value}{/3} define the
name of a monitored value, the module's name and a unique key. The second can
be either an \code{rrd()} or an update fun.
The \erlfun{monitor}{proc\_check\_timeslot}{/3} function can be used if your
module does not regularly create new data. In this case, the monitor process
would not have the latest data for others to retrieve. This function forces a
check and creates the new time slot if needed (thus reporting the data).

This is how forwarding works (taken from \erlmodule{api\_tx}):

\begin{lstlisting}
monitor:client_monitor_set_value(
  ?MODULE, 'req_list',
  fun(Old) ->
    Old2 = case Old of
             % 10s monitoring interval, only keep newest in the client process
             undefined -> rrd:create(10 * 1000000, 1, {timing, ms});
             _ -> Old
           end,
    rrd:add_now(TimeInUs / 1000, Old2)
  end),
\end{lstlisting}

As in this case there is no safe way of initialising the value, it is more
useful to provide an update fun to
\erlfun{monitor}{client\_monitor\_set\_value}{/3}. This function is only useful
for the client processes as it reports to the monitor in the
\code{clients_group} (recall that client processes do not belong to any group).
All other processes should use \erlfun{monitor}{monitor\_set\_value}{/3} with
the same semantics.

\section{Writing Unit Tests}

\subsection{Plain Unit Tests}

\subsection{Randomized Testing Using \texttt{tester}}

\subsection{Randomized Testing Using \texttt{proto\_sched}}
